---
title: "Análisis descriptivo en pacientes diagnosticados de COVID-19 en el Hospital de Especialidades IMSS Puebla"
author: "Luis G. Vázquez de Lara Cisneros."
date: "27/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
```

```{r muestra, message=FALSE, warning=FALSE}
load('Rdata/bdcoving.rda')
muestra <- dim(bdcoving)[1]
fechaini <- min(bdcoving$fecha)
fechafin <- max(bdcoving$fecha)
covpos <- sum(bdcoving$rescov1 == 1)
noegre <- sum(!(bdcoving$motivoegre %in% c(2,3)))
totmuestra <- covpos - noegre
```

## Características de la muestra
This study was conducted in the  _Unidad Médica de Alta Especialidad (UMAE) Hospital de Especialidades Gral. Div. Manuel Ávila Camacho_, Puebla, Puebla, Mexico.
A total of  ``r muestra`` hospital records between ``r fechaini`` and ``r fechafin`` with diagnosis of covid-19 were analyzed, ``r covpos`` had a positive result on RT-PCR of a nasopharyngeal swab, and a total of ``r noegre`` patients were transferred to other facilities. The final sample comprised ``r totmuestra`` hospital records.

## Comorbidities

```{r frec-app, message=FALSE, warning=FALSE}
app <- read_lines('bases-originales/nomapp.txt')
app <- str_split(app, '\t')[[1]]%>%str_replace_all('\\s|-', "_") 
bdcoving <- bdcoving %>%
  mutate(app_18 = as.factor(ifelse(imc >= 25 & imc < 30, 1, 2)),
         app_19 = as.factor(ifelse(imc >= 30, 1, 2)))
app <- c('HAS', app, 'SOBREPESO', 'OBESIDAD')
nomappord <- bdcoving%>%select(starts_with('app'))%>%names

frecapp <- bdcoving%>%
  filter(motivoegre %in% c(2,3), rescov1 == 1)%>% 
  pivot_longer(starts_with('app'), names_to = 'apps', values_to = 'valor')%>%
  group_by(apps, valor) %>%
    mutate(apps = factor(apps, levels = nomappord, labels = app),
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No')))%>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%filter(valor == 'Si')%>%
  arrange(frec)%>%
  rename(Antecedentes = apps, Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Antecedentes, Cantidad, Porcentaje, Total)
kable(frecapp)
```

## Association between comorbidities and outcome

```{r frecapp-egreso}
bdcat <- bdcoving%>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, starts_with('app')))
listachis <- apply(bdcat[,-1],2, function(i){chisq.test(table(i, bdcat$motivoegre))$p.value})
dfchis <- data.frame(apps = app, p = round(listachis, 3))

frecapp_egreso <- bdcoving%>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>%
  select(c(motivoegre, starts_with('app'))) %>%
  pivot_longer(starts_with('app'), names_to = 'apps', values_to = 'valor')%>%
  group_by(motivoegre, apps, valor) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')),
         apps = factor(apps, levels = nomappord, labels = app),
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No')))%>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  left_join(dfchis, by = 'apps' ) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos, p) %>%
  arrange(p, by_group = TRUE)%>%
  filter(apps != 'CONJUNTIVITIS') %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)
kable(frecapp_egreso)


```

## Information of combined comorbidities

```{r numapp-glob, message=FALSE, warning=FALSE}
facanum <- function(x){
  as.numeric(as.character(x))
}

numappglob <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1)%>% 
  rowwise()%>%
  mutate(across(starts_with('app'), facanum) ) %>%
  mutate(numapp = (length(nomappord)-sum(is.na(c_across(starts_with('app')))))*2 - sum(c_across(starts_with('app')), na.rm = TRUE))%>%
  select(motivoegre, numapp)%>%
  ungroup %>%
  mutate(numapp = case_when(
    numapp == 0 ~ 'No',
    numapp == 1 ~ 'Una',
    numapp > 1  ~ 'Más de una'
  ))%>%
  mutate(numapp = factor(numapp, levels = c('No', 'Una', 'Más de una')))%>%
  group_by(numapp)%>%
    summarise(n = n()) %>%
    mutate(frec = round(n/sum(n)*100, 2),
           total = sum(n),
           validos = sum(n)- sum(is.na(numapp)))%>%
    select(numapp, n, frec, validos) %>%
    rename(`Número de enfermedades previas` = numapp,
           Número = n,
           Porcentaje = frec,
           Total = validos)
kable(numappglob)
```

## Association of number of combined comorbidities and outcome.

```{r label = numapp-motivoegre}
facanum <- function(x){
  as.numeric(as.character(x))
}
numappchi <- bdcoving %>%
  filter(motivoegre %in% c(2,3)  & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  rowwise()%>%
  mutate(across(starts_with('app'), facanum) ) %>%
  mutate(numapp = (length(nomappord)-sum(is.na(c_across(starts_with('app')))))*2 - sum(c_across(starts_with('app')), na.rm = TRUE))%>%
  select(motivoegre, numapp)%>%
  ungroup %>%
  mutate(numapp = case_when(
    numapp == 0 ~ 'No',
    numapp == 1 ~ 'Una',
    numapp > 1  ~ 'Más de una'
  ))%>%
  mutate(numapp = factor(numapp, levels = c('No', 'Una', 'Más de una')))

tablanumapp <- table(`Número de enfermedades previas` = numappchi$numapp, Evolución = numappchi$motivoegre)
tablapropnumapp <- round(prop.table(tablanumapp, margin = 2)*100,2)
chinumapp <- chisq.test(table(numappchi$numapp, numappchi$motivoegre))$p.value
chinumapp <- round(chinumapp, 3)

numapps_egreso <-numappchi %>%
  group_by(motivoegre, numapp)%>%
  summarise(n = n()) %>%
  pivot_wider(names_from = motivoegre, values_from = n) %>%
  mutate(total = `Defunción` + `Mejoría`,
         pordef = round(`Defunción`/total*100, 1),
         pormejo = round(`Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`Mejoría`,4,side = 'right'), '(', pormejo, '%)'),
         p = c(chinumapp[1], '', '')) %>% 
  select(numapp, total, muertos, vivos, p) %>%
  #arrange(p, by_group = TRUE)%>%
  rename(`Number of associated comorbidities` = numapp, Total = total, `Defunción` = muertos, `Mejoría` = vivos)
kable(numapps_egreso)
```

## Treatments before hospitalization.
```{r tx-prev }
txprev <- read_lines('bases-originales/nomtxprev.txt')
txprev <- str_split(txprev, '\t')[[1]] %>%
  str_trim()%>%
  str_replace_all('\\s|-', "_")
nomtxprevord <- bdcoving%>%select(starts_with('txprev'))%>%names

frectxprev <- bdcoving%>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  pivot_longer(starts_with('txprev'), names_to = 'txprevs', values_to = 'valor')%>%
  group_by(txprevs, valor) %>%
  mutate(txprevs = factor(txprevs, levels = nomtxprevord, labels = txprev), 
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No')))%>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%filter(valor == 'Si')%>%
  arrange(desc(frec))%>%
  rename(Antecedentes = txprevs, Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Antecedentes, Cantidad, Porcentaje, Total)
kable(frectxprev)

```

## Treatments before hospitalization by outcome

```{r txprev-motivoegre}
bdcat <- bdcoving%>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, starts_with('txprev')))
listachis <- apply(bdcat[,2:23],2, function(i){chisq.test(table(i, bdcat$motivoegre))$p.value})

dfchis <- data.frame(apps = txprev, p = round(listachis, 3))

frectxprev_egreso <- bdcoving%>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>%
  select(c(motivoegre, starts_with('txprev'))) %>%
  pivot_longer(starts_with('txprev'), names_to = 'apps', values_to = 'valor')%>%
  group_by(motivoegre, apps, valor) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')),
         apps = factor(apps, levels = nomtxprevord, labels = txprev),
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No')))%>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  left_join(dfchis, by = 'apps' ) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos, p) %>%
  arrange(p, by_group = TRUE)%>%
  filter(!(apps %in% c('DIÓXIDO_DE_CLORO', 'INTERFERON', 'ANAFERON' ))) %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)
kable(frectxprev_egreso)

```

## Number of treatment combinations before hospitalization 

```{r numtxprev-global}
facanum <- function(x){
  as.numeric(as.character(x))
}

numtxprevglob <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  mutate(across(starts_with('txprev'), facanum) ) %>%
  rowwise()%>%
  mutate(numtxprev = (length(nomtxprevord)-sum(is.na(c_across(starts_with('txprev')))))*2 - sum(c_across(starts_with('txprev')), na.rm = TRUE))%>%
  select(motivoegre, numtxprev) %>%
  ungroup %>%
  mutate(numtxprev = case_when(
    numtxprev == 0 ~ 'No',
    numtxprev == 1 ~ 'Uno',
    numtxprev > 1  ~ 'Más de uno'
  ))%>%
  mutate(numtxprev = factor(numtxprev, levels = c('No', 'Uno', 'Más de uno')))%>%
  group_by(numtxprev)%>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n),
         validos = sum(n)- sum(is.na(numtxprev)))%>%
  select(numtxprev, n, frec, validos) %>%
  rename(`Número de tratamientos previos` = numtxprev,
         Número = n,
         Porcentaje = frec,
         Total = validos)
kable(numtxprevglob)
```

## Number of treatment combinations before hospitalization by outcome

```{r numtxprev-motivoegre}
facanum <- function(x){
  as.numeric(as.character(x))
}
numtxprevchi <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  rowwise()%>%
  mutate(across(starts_with('txprev'), facanum) ) %>%
  mutate(numtxprev = (length(nomtxprevord)-sum(is.na(c_across(starts_with('txprev')))))*2 - sum(c_across(starts_with('txprev')), na.rm = TRUE))%>%
  select(motivoegre, numtxprev)%>%
  ungroup %>%
  mutate(numtxprev = case_when(
    numtxprev == 0 ~ 'No',
    numtxprev == 1 ~ 'Uno',
    numtxprev > 1  ~ 'Más de uno'
  ))%>%
  mutate(numtxprev = factor(numtxprev, levels = c('No', 'Uno', 'Más de uno')))

chinumtxprev <- chisq.test(table(numtxprevchi$numtxprev, numtxprevchi$motivoegre))$p.value
chinumtxprev <- round(chinumtxprev, 3)

numtxprevs <-numtxprevchi %>% group_by(motivoegre, numtxprev)%>%
  summarise(n = n()) %>%
  pivot_wider(names_from = motivoegre, values_from = n) %>%
  mutate(total = `Defunción` + `Mejoría`,
         pordef = round(`Defunción`/total*100, 1),
         pormejo = round(`Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`Mejoría`,4,side = 'right'), '(', pormejo, '%)'),
         p = c(chinumtxprev, '', '')) %>% 
  select(numtxprev, total, muertos, vivos, p) %>%
  rename(`Número de tratamientos previos` = numtxprev,
         Total =total,
         `Defunción` = muertos,
         `Mejoría` = vivos)
kable(numtxprevs)
```

## Treatments during hospitalization

```{r frectxhosp-global}
txhosp <- read_lines('bases-originales/nomtxhosp.txt')
txhosp <- str_split(txhosp, '\t')[[1]] %>%
  str_trim()%>%
  str_replace_all('\\s|-', "_") 
nomtxhospord <- bdcoving%>%select(starts_with('txhosp'))%>%names

frectxhosp <- bdcoving%>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  pivot_longer(starts_with('txhosp'), names_to = 'txhosps', values_to = 'valor')%>%
  group_by(txhosps, valor) %>%
  mutate(txhosps = factor(txhosps, levels = nomtxhospord, labels = txhosp),
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No')))%>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%filter(valor == 'Si')%>%
  arrange(desc(frec))%>%
  rename(Tratamientos = txhosps, Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Tratamientos, Cantidad, Porcentaje, Total)
kable(frectxhosp)

```

## Treatments during hospitalization by outcome

```{r frectxhosp-egreso}
bdcat <- bdcoving%>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, starts_with('txhosp')))
listachis <- apply(bdcat[,2:21],2, function(i){chisq.test(table(i, bdcat$motivoegre))$p.value})

dfchis <- data.frame(apps = txhosp, p = round(listachis, 3))

frectxhosp_egreso <- bdcoving%>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>%
  select(c(motivoegre, starts_with('txhosp'))) %>%
  pivot_longer(starts_with('txhosp'), names_to = 'apps', values_to = 'valor')%>%
  group_by(motivoegre, apps, valor) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')),
         apps = factor(apps, levels = nomtxhospord, labels = txhosp),
         valor = factor(valor, levels = c(1,2), labels = c('Si', 'No'))) %>%
  summarise(n = n())%>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  left_join(dfchis, by = 'apps' ) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos, p) %>%
  arrange(p, by_group = TRUE)%>%
  rename(`Tratamientos previos` = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)
kable(frectxhosp_egreso)

```

## Number of treatment combinations during hospitalization

```{r numtxhosp-global}
facanum <- function(x){
  as.numeric(as.character(x))
}

numtxhospglob <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  mutate(across(starts_with('txhosp'), facanum) ) %>%
  rowwise()%>%
  mutate(numtxhosp = (length(nomtxhospord)-sum(is.na(c_across(starts_with('txhosp')))))*2 - sum(c_across(starts_with('txhosp')), na.rm = TRUE))%>%
  select(motivoegre, numtxhosp) %>%
  ungroup %>%
  group_by(numtxhosp)%>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n),
         validos = sum(n)- sum(is.na(numtxhosp)))%>%
  select(numtxhosp, n, frec, validos) %>%
  rename(`Número de fármacos administrados` = numtxhosp,
         Número = n,
         Porcentaje = frec,
         Total = validos)
kable(numtxhospglob)
```

## Combination of treatments during hospitalization

```{r combtx-hosp}
fcomb <- function(x){
  cad <- str_c(x, sep = ',', collapse = TRUE)
  return(cad)
}
#cargo nobres abreviados de los tx hospitalarios:
txhospabr <- read_lines('bases-originales/nomtxhospabr.txt')
txhospabr <- str_split(txhospabr, '\t')[[1]] %>%
  str_trim()%>%
  str_replace_all('\\s|-', "_") 

combtxhosp <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>%
  select(c(nss, motivoegre, starts_with('txhosp'))) %>%
  rename_with( ~ txhospabr, starts_with('txhosp')) %>%
  pivot_longer(c(3:22), names_to = 'txhosps', values_to = 'valor') %>%
  filter(valor == 1)%>%
  group_by(nss, motivoegre) %>%
  arrange(txhosps) %>%
  summarise(comb = str_c(txhosps,  collapse = ',')) %>% ungroup() %>%
  group_by(comb) %>%
  summarise(n = n(),
            mort = sum(motivoegre == 3),
            propmort = round(100*sum(motivoegre == 3)/n(), 1)) %>%
  arrange(desc(n))# %>%filter(n > 1)
kable(combtxhosp)

```

# Colchicine plus clopidogrel treatment.

```{r comb-cochiclopi}
comb2 <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(nss, motivoegre, starts_with('txhosp'))) %>%
  #mutate(across(starts_with('txhosp'), facanum)) %>%
  mutate(colchiclopi = case_when(
    txhosp_9 == 1 & txhosp_13 == 1 ~ 'colchi+clopi',
    txhosp_9 == 1 & txhosp_13 == 2 ~ 'colchicina',
    txhosp_9 == 2 & txhosp_13 == 1 ~ 'clopidogrel',
    TRUE ~ 'otros'
  )) %>%
  select(nss, motivoegre, colchiclopi)
  
tcomb2 <- table(outcome = comb2$motivoegre, Treatment = comb2$colchiclopi)
kable(tcomb2)
chisq.test(tcomb2)

```

# Mortality in treatments which included colchicine

```{r combtx-concolchi}
bdcolchis <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>%
  select(c(nss, motivoegre, starts_with('txhosp'))) %>%
  rename_with( ~ txhospabr, starts_with('txhosp')) %>%
  rename(aacolchi = colchi) %>%
  pivot_longer(c(3:22), names_to = 'txhosps', values_to = 'valor') %>%
  filter(valor == 1)%>%
  group_by(nss, motivoegre) %>%
  arrange(txhosps) %>%
  summarise(comb = str_c(txhosps,  collapse = ',')) %>% ungroup() %>%
  group_by(comb) %>%
  summarise(n = n(),
            mort = sum(motivoegre == 3),
            propmort = round(100*sum(motivoegre == 3)/n(), 1)) %>%
  arrange(desc(n)) %>%
  slice(str_which(comb, 'colchi')) %>%
  arrange(propmort)
kable(bdcolchis)
```

## Demographic information

```{r demographics-num}
bdvarnumglob <- bdcoving %>%filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  select(edad, peso, talla, imc)%>%
  summarise(across(everything(), list(
    Promedio = ~ round(mean(.x, na.rm = TRUE),1),
    `Desviación estándar` = ~ round(sd(.x, na.rm = TRUE),1)
  )))%>%
  pivot_longer(cols = everything(),
               names_sep = '_',
               names_to = c('Variable', '.value'))%>%
  unite('Promedio±DE', c(Promedio, `Desviación estándar`), sep = ' ± ')
kable(bdvarnumglob, caption = 'Numeric variables')
```

```{r demographics-cat}
frecsexo <- bdcoving%>%
  filter(motivoegre %in% c(2,3), rescov1 == 1)%>%
  select(sexo)%>%
 pivot_longer(sexo, names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(frec)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)

frecocup <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1, ocupacion != 7)%>%
  select(ocupacion)%>%
  mutate(ocupacion = case_when(
    ocupacion == 1 ~ 'Personal sanitario',
    ocupacion == 2 ~ 'Trabajo de oficina',
    ocupacion == 3 ~ 'Trabajo al aire libre',
    ocupacion == 4 ~ 'Trabajo en espacio público',
    ocupacion == 5 ~ 'Trabajo en casa',
    ocupacion == 6 ~ 'No trabaja'
  ))%>%
  rename( `Ocupación` = ocupacion) %>%
  pivot_longer(`Ocupación`, names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(frec)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)

frecescol <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1)%>%
  select(escolaridad)%>%
  mutate(escolaridad = case_when(
    escolaridad == 1 ~ 'Analfabeta',
    escolaridad == 2 ~ 'Primaria',
    escolaridad == 3 ~ 'Secundaria',
    escolaridad == 4 ~ 'Bachillerato',
    escolaridad == 5 ~ 'Licenciatura',
    escolaridad == 6 ~ 'Posgrado'
  ))%>%
  rename( Escolaridad = escolaridad) %>%
  na.omit %>%
  pivot_longer(Escolaridad, names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(frec)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)

frecnivsoc <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1)%>%
  select(nivsoc)%>%
  mutate(nivsoc = case_when(
    nivsoc %in% c(1,2) ~ 'Bajo, medio-bajo',
    nivsoc %in% c(3,4) ~ 'Medio-alto, alto'
  ))%>%
  na.omit %>%
  rename( `Nivel socioeconómico` = nivsoc) %>%
  pivot_longer(`Nivel socioeconómico`, names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(frec)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)

dbvarsocdemglob <- bind_rows(frecsexo,frecocup,frecescol,frecnivsoc) %>%
  mutate (`n (%)` = paste0(Cantidad, ' (', Porcentaje, ')')) %>%
  select(Variable, Valor, `n (%)`, Total)
kable(dbvarsocdemglob, caption = 'Categorical variables')

```

## Demographic information by outcome

```{r demographics-num-motivoegre}
bdnum <- bdcoving%>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, edad, peso, talla, imc))
listates <- apply(bdnum[,-1],2, function(i){t.test(i ~ bdnum$motivoegre)$p.value})
dftes <- data.frame(Variable = c('edad', 'peso', 'talla', 'imc'), p = round(listates, 3))

bdvarnummotiegre <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(motivoegre, edad, peso, talla, imc)%>%
  group_by(motivoegre) %>%
  summarise(across(everything(), list(
    Promedio = ~ round(mean(.x, na.rm = TRUE),2),
    `Desviación estándar` = ~ round(sd(.x, na.rm = TRUE),2)
  )))%>%
  pivot_longer(cols = -motivoegre,
               names_sep = '_',
               names_to = c('Variable', '.value')) %>%
  left_join(dftes, by = 'Variable')%>%
  unite('Promedio±DE', c(Promedio, `Desviación estándar`), sep = ' ± ') %>%
  select(Variable, motivoegre, `Promedio±DE`, p  )%>%
  rename(`Motivo de egreso` = motivoegre)%>%
  arrange(Variable)
kable(bdvarnummotiegre, caption = 'Numeric variables')
```

```{r demographics-cat-motivoegre}
frecsexoegre <- bdcoving%>%
  filter(motivoegre %in% c(2,3), rescov1 == 1)%>%
  select(motivoegre, sexo)%>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción'))) %>%
  pivot_longer(sexo, names_to = 'apps', values_to = 'valor') %>%
  group_by(apps,motivoegre, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos) %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)

frecocupegre <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1, ocupacion != 7)%>%
  select(motivoegre, ocupacion)%>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción'))) %>%
  mutate(ocupacion = case_when(
    ocupacion == 1 ~ 'Personal sanitario',
    ocupacion == 2 ~ 'Trabajo de oficina',
    ocupacion == 3 ~ 'Trabajo al aire libre',
    ocupacion == 4 ~ 'Trabajo en espacio público',
    ocupacion == 5 ~ 'Trabajo en casa',
    ocupacion == 6 ~ 'No trabaja'
  ))%>%
  rename( `Ocupación` = ocupacion) %>%
  pivot_longer(`Ocupación`, names_to = 'apps', values_to = 'valor') %>%
  group_by(motivoegre, apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>%
  select(apps, valor, total, muertos, vivos) %>%
  filter(!is.na(valor)) %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)

frecescolegre <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1)%>%
  select(motivoegre, escolaridad)%>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción'))) %>%
  mutate(escolaridad = case_when(
    escolaridad == 1 ~ 'Analfabeta',
    escolaridad == 2 ~ 'Primaria',
    escolaridad == 3 ~ 'Secundaria',
    escolaridad == 4 ~ 'Bachillerato',
    escolaridad == 5 ~ 'Licenciatura',
    escolaridad == 6 ~ 'Posgrado'
  ))%>%
  rename( Escolaridad = escolaridad) %>%
  na.omit %>%
  pivot_longer(Escolaridad, names_to = 'apps', values_to = 'valor') %>%
  group_by(motivoegre, apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% #usé unite con las numéricas.
  select(apps, valor, total, muertos, vivos) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos) %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)

frecnivsocegre <- bdcoving%>%
  filter(motivoegre %in% c(2,3),rescov1 == 1)%>%
  select(motivoegre, nivsoc)%>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción'))) %>%
  mutate(nivsoc = case_when(
    nivsoc %in% c(1,2) ~ 'Bajo, medio-bajo',
    nivsoc %in% c(3,4) ~ 'Medio-alto, alto'
  ))%>%
  rename( `Nivel socioeconómico` = nivsoc) %>%
  pivot_longer(`Nivel socioeconómico`, names_to = 'apps', values_to = 'valor') %>%
  group_by(motivoegre, apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2),
         total = sum(n)) %>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  filter(!is.na(valor) )%>%
  select(apps, valor, total, muertos, vivos) %>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)

dbvarsocdemegre <- bind_rows(frecsexoegre,frecocupegre,frecescolegre,frecnivsocegre)
#______________________________________________________________________________________
fchipasocdem <- function(x, y){
  #p <- round(fisher.test(table(x, y), simulate.p.value = TRUE)$p.value, 3)
  p <- round(chisq.test(table(x, y))$p.value, 3)
  if (!is.factor(x)){
    x <- as.factor(x)
  } 
  blancos <- length(levels(x)) -1
  if (p < 0.001){
    p <- '< 0.001'
  }
  cad <- c(p, rep('', blancos))
  return(cad)
}

bdsexo <-bdcoving %>%
  filter(motivoegre %in% c(2,3), rescov1 == 1) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3)))%>%
  select(motivoegre, sexo)
chisexo <-   fchipasocdem(bdsexo$sexo, bdsexo$motivoegre)

bdocup <- bdcoving %>%
  filter(motivoegre %in% c(2,3), rescov1 == 1, ocupacion != 7) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3)),
         ocupacion = factor(ocupacion, levels = c(1,2,3,4,5,6)))%>%
  select(motivoegre, ocupacion)
chiocup <- fchipasocdem(bdocup$ocupacion, bdocup$motivoegre)

bdescol <- bdcoving %>%
  filter(motivoegre %in% c(2,3), rescov1 == 1) %>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3)),
         escolaridad = factor(escolaridad, levels = c(1,2,3,4,5,6)))%>%
  select(motivoegre, escolaridad)
chiescol <- fchipasocdem(bdescol$escolaridad, bdescol$motivoegre)

bdnivsoc <- bdcoving %>%
  filter(motivoegre %in% c(2,3), rescov1 == 1) %>%
  mutate(nivsoc = case_when(
    nivsoc %in% c(1,2) ~ 'Bajo, medio-bajo',
    nivsoc %in% c(3,4) ~ 'Medio-alto, alto'
  ))%>%
  mutate(motivoegre = factor(motivoegre, levels = c(2,3)),
         nivsoc = as.factor(nivsoc))%>%
  select(motivoegre, nivsoc)
chinivsoc <- fchipasocdem(bdnivsoc$nivsoc, bdnivsoc$motivoegre)

chisvarsocdem <- c(chisexo, chiocup, chiescol, chinivsoc)

dbvarsocdemegre <- dbvarsocdemegre %>%
  ungroup %>%
  mutate(p = chisvarsocdem)
kable(dbvarsocdemegre, caption = 'Categorial variables')

```

### Number of patients with a second test for SARS-Cov2
```{r}
bdcov2 <- bdcoving %>%
  mutate(cov2 = factor(cov2, levels = c(1,2), labels = c('Si', 'No'))) %>%
  select(cov2)

kable(table(bdcov2$cov2), col.names = c('Segunda prueba', 'Número'))

```

```{r variable-varclincat}

varclincat <- bdcoving%>%
  select(c(84:157, 159)) %>%
  select(where(is.factor))%>%
  names
```

## Análisis previo de las variables clínicas categóricas

```{r varclin-bruta}

frecvclinglob <- bdcoving%>%
  filter(motivoegre %in% c(2,3), rescov1 == 1)%>%
  pivot_longer(all_of(varclincat),names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(apps)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)

bdvarclin <- bdcoving %>%
  filter(motivoegre %in% c(2,3), rescov1 == 1)%>%
  select(c(all_of(varclincat), motivoegre))%>%
  mutate(across(where(is.factor), as.character)) %>%
  mutate(txhashosp = case_when(txhashosp == '1' ~ 'Diurético',
                               txhashosp == '1,2,5,7' ~ 'Diur,araII,baldos,calcioant',
                               txhashosp == '2' ~ 'ARAII',
                               txhashosp == '2,4' ~ 'ARAII',
                               txhashosp == '3' ~ 'IECA',
                               txhashosp == '3, 7' ~ 'ARAII,calcioant',
                               txhashosp == '4' ~ 'Betabloqueador',
                               txhashosp == '6' ~ 'Sin medicamento',
                               txhashosp %in% c('7','8,7') ~ 'Calcioantagonista',
                               TRUE ~ txhashosp),
         anticoag = case_when(anticoag == '1' ~ 'Profilaxis',
                              anticoag == '2' ~ 'Anticoagulante',
                              TRUE ~ NA_character_),
         controltahosp = case_when(controltahosp == '1' ~ 'Si',
                                   controltahosp == '2' ~ 'No'),
         dislip = ifelse(dislip == '1', 'Si', 'No'),
         neumonia = ifelse(neumonia == '1', 'Si', 'No'),
         piel = ifelse(piel == '1', 'Si', 'No'),
         sintoma1 = case_when(sintoma1 == '1' ~ 'Tos',
                              sintoma1 == '2' ~ 'Fiebre',
                              sintoma1 == '3' ~ 'Cefalea',
                              sintoma1 == '4' ~ 'Anosmia',
                              sintoma1 == '5' ~ 'Malestar general',
                              sintoma1 == '6' ~ 'Mareo',
                              sintoma1 == '7' ~ 'Debilidad muscular',
                              sintoma1 == '8' ~ 'Diarrea',
                              sintoma1 == '9' ~ 'Pérdida gusto',
                              sintoma1 == '10' ~ 'Erupciones cutáneas',
                              sintoma1 == '11' ~ 'Disnea',
                              sintoma1 == '12' ~ 'Dolor u opresión torácica',
                              sintoma1 == '13' ~ 'Incapacidad',),
         tipoing = case_when(tipoing == 1 ~ 'Piso',
                             tipoing == 2 ~ 'UCI', 
                             tipoing == 3 ~ 'Urgencias'),
         ventilador = case_when(ventilador == 1 ~ 'Si',
                                ventilador == 2 ~ 'No',
                                TRUE ~ NA_character_),
         ventnoinv = case_when(ventnoinv == 1 ~ 'Si',
                               ventnoinv == 2 ~ 'No',
                               TRUE ~ NA_character_),
         motivoegre = ifelse(motivoegre == 2, 'Mejoría', 'Defunción'))

frecvclin2 <- bdvarclin %>% pivot_longer(all_of(varclincat),names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(apps)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)
kable(frecvclin2, caption = 'Datos brutos')
```


## Clinical characteristics 

```{r frecvarlin-catglob}
frecvclin3 <- bdvarclin %>%
  pivot_longer(c(-starts_with(c('ekg', 'gdo')), -txhashosp, -piel, -ventnoinv, -neumonia),names_to = 'apps', values_to = 'valor') %>%
  group_by(apps, valor) %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  arrange(apps)%>%
  rename(Variable = apps,Valor = valor , Cantidad = n, Porcentaje = frec, Total = total)%>%
  select(Variable, Valor, Cantidad, Porcentaje, Total)
kable(frecvclin3, caption = 'Variables seleccionadas')
```


```{r varclin-numglob}
nomvar <- bdcoving %>% select(where(is.numeric)) %>% names
nomvar <- nomvar[-c(1:5, 39:40, 52)]
dimbdcoving <- dim(bdcoving %>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% select(nss))[1]
validos <- function(x){
  dimbdcoving - sum(is.na(x))
}

bdvparaclinglob <- bdcoving %>%filter(motivoegre %in% c(2,3) & rescov1 == 1)%>%
  select(all_of(nomvar))%>%
  summarise(across(everything(), list(
    Validos = ~ validos(.x),
    Promedio = ~ round(mean(.x, na.rm = TRUE),1),
    `Desviación estándar` = ~ round(sd(.x, na.rm = TRUE),1)
  )))%>%
  pivot_longer(cols = everything(),
               names_sep = '_',
               names_to = c('Variable', '.value'))%>%
  unite('Promedio ± DE', c(Promedio, `Desviación estándar`), sep = ' ± ')
kable(bdvparaclinglob, caption = 'Initial laboratory measures')
```

## Clinical characteristics by outcome

```{r varclin-cat-motivoegre}


nompachisvclin <- bdvarclin %>%
  select(c(-starts_with(c('ekg', 'gdo')), -txhashosp, -piel, -ventnoinv, -neumonia, -motivoegre)) %>% names
listachisvclin <- apply(bdvarclin[, nompachisvclin],2, function(i){chisq.test(table(i, bdvarclin$motivoegre))$p.value})
listachisvclin <- round(listachisvclin, 3)
listachisvclin <- data.frame(apps = nompachisvclin, p = listachisvclin)

bdvarclinred_nonas <- bdvarclin %>%
  select(c(all_of(nompachisvclin), motivoegre)) %>%
  mutate(sintoma1 = ifelse(sintoma1 %in% c('Debilidad muscular', 'Incapacidad','Diarrea'), 'Otros',   sintoma1)) %>%
  pivot_longer(1:6, names_to = 'apps', values_to = 'valor' ) %>%
  group_by(motivoegre, apps, valor) %>%
  na.omit %>%
  summarise(n = n()) %>%
  mutate(frec = round(n/sum(n)*100, 2), total = sum(n))%>%
  pivot_wider(names_from = motivoegre, values_from = c(n, frec, total)) %>%
  mutate(total = `n_Defunción` + `n_Mejoría`,
         pordef = round(`n_Defunción`/total*100, 1),
         pormejo = round(`n_Mejoría`/total*100, 1),
         muertos = paste0(str_pad(`n_Defunción`,4,side = 'right', pad = ' '), '(', pordef, '%)'),
         vivos = paste0(str_pad(`n_Mejoría`,4,side = 'right'), '(', pormejo, '%)')) %>% 
  select(apps, valor, total, muertos, vivos) %>%
  left_join(listachisvclin, by = 'apps') %>%
  arrange(p)%>%
  rename(Variable = apps,Valor = valor , Total = total, `Defunción` = muertos, `Mejoría` = vivos)

kable(bdvarclinred_nonas, caption = 'Categorical variables')

```


```{r varclin-num-motivoegre}
nomvar2 <- bdvparaclinglob %>%
  filter(Validos > 100) %>%
  select(Variable) %>% pull

bdnum2 <- bdcoving%>%filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, all_of(nomvar2)))
listates2 <- apply(bdnum2[,-1],2, function(i){t.test(i ~ bdnum2$motivoegre)$p.value})
dftes2 <- data.frame(Variable = nomvar2, p = round(listates2, 3))

validos2 <- function(x){
  length(x) - sum(is.na(x))
}

bdvarnum2motiegre <- bdcoving %>%
  filter(motivoegre %in% c(2,3) & rescov1 == 1) %>% 
  mutate(motivoegre = factor(motivoegre, levels = c(2,3), labels = c('Mejoría', 'Defunción')))%>%
  select(c(motivoegre, all_of(nomvar2)))%>%
  group_by(motivoegre) %>%
  summarise(across(everything(), list(
    Validos = ~ validos2(.x),
    Promedio = ~ round(mean(.x, na.rm = TRUE),2),
    `Desviación estándar` = ~ round(sd(.x, na.rm = TRUE),2)
  )))%>%
  pivot_longer(cols = -motivoegre,
               names_sep = '_',
               names_to = c('Variable', '.value')) %>%
  left_join(dftes2, by = 'Variable')%>%
  unite('Promedio ± DE', c(Promedio, `Desviación estándar`), sep = ' ± ') %>%
  select(Variable, Validos, motivoegre, `Promedio ± DE`, p  )%>%
  rename(`Motivo de egreso` = motivoegre)%>%
  arrange(p, Variable)
kable(bdvarnum2motiegre, caption = 'Numeric variables')
```